---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

The following is an exploration and dive into the Behavioral Risk Factor Surveillance System (BRFSS) data, which is built of a collection of 400,000 plus annual interveiws with questions and answers regarding a wide range of health related characteristics. 

### Load packages

The following are packages needed to run the below analysis of the BRFSS data.

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

The data is read in via a url link to the dataset. 

```{r load-data, cache=TRUE}
URL <- "https://d18ky98rnyall9.cloudfront.net/_384b2d9eda4b29131fb681b243a7767d_brfss2013.RData?Expires=1487808000&Signature=G8IePZUxnzkO82tsaOF2jVaSuIHVPI9xz-SHy8TK-EK2FQjB5cqjzGFZGcHmzqN2jVPYxOMtBhoxMk3sDSBzzn3dBRYoFAwyJltmUdKltcF0hvXyArEaO~k8GhQ5~AsT1ZcFzQh7S-uEHNH6XG50YuSR~UCt~9noa0WbnSTbUNY_&Key-Pair-Id=APKAJLTNE6QMUY6HBC5A"
download.file(URL, "~/Desktop/Coursera/Intro to Stats/Project/brfss2013.RData", method = "curl")
load("~/Desktop/Coursera/Intro to Stats/Project/brfss2013.RData")
```

* * *

## Part 1: Data

The BRFSS data is "observational" rather than "experimental" in nature, which immediately rules out any conclusions that can be drawn regarding causal relationships between variables. This type of data does, however, enable relationships to be generalized to the total population the data was sampled from. That being said, the data must have been collected using random sampling in order for it to be truely generalizable. 

The data gathered via the BRFSS is done so "randomly", but the data is still collected via telephone interveiws strictly. This introduces a "convenience sampling" problem because those that are interveiwed are only those with a telephone (either land line or cellphone). 

When collecting data over the phone, certain variables are also likely to be influenced by the medium of the phone. Also, those interveiwed may show a systematic reluctance to share truthful information regarding certain questions. This would be reflective of the self-serving bias. 

* * *

## Part 2: Research questions

**Research quesion 1:**

How does the amount of sleep reported relate to overall health reported?

**Research quesion 2:**

How does the amount of exercise reported relate to the amount of sleep reported?

**Research quesion 3:**

How does income relate to overall health reported?

* * *

## Part 3: Exploratory data analysis

Below each questions listed above will be further explored and the given data will be processed in order to allow this data analysis. 

**Research quesion 1:**

Initially we take a look at raw data, ultimately hoping to get a look at the summary statistics and distribution of the amount of sleep reported. Below the "NA" or missing values are removed and then a histogram showing the amount of sleep reported across all those interveiwed is coded. 

```{r All Sleep Processing}
sleep <- brfss2013$sleptim1[!(is.na(brfss2013$sleptim1))]
ggplot() + aes(sleep)+ geom_histogram(binwidth=1, colour="black", fill="white")
```

The histogram reveals that there are outliers within the sleep data. We can explore these outliers in order to determine which values should be included, and which should not.

```{r Sleep Outliers}
unique(sleep)
```

Because no one is capable of not sleeping and surviving over any significant amount of time and sleeping greater than 24 hours is impossible, as the day is only 24 hours long, some of the collected data can be removed. We will now eliminate all values that are 0 or 24 and greater, then take another look at the distribution of data.

```{r Cleaned Sleep Histo}
sleep <- subset(sleep, sleep > 0 & sleep < 24)
ggplot() + aes(sleep)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = 7.05, color = "blue", linetype = 2)
summary(sleep)
```

While there are still values very close to 24 hours and 0 hours of sleep, whether these are accurate data points is up for some debate, and so they will be included with the cleaned sleep data. We can see that the mean and median are both very close to 7 hours of sleep across the population.

Now that we have the sleep data relatively clean, we can process the general health data according to the amount of sleep each participant recorded. 

```{r Excellent Gen Health and Sleep}
Excellent <- subset(brfss2013, brfss2013$genhlth == "Excellent")
Excellent <- Excellent$sleptim1[!(is.na(Excellent$sleptim1))]
Excellent <- subset(Excellent, Excellent > 0 & Excellent < 24)
ggplot() + aes(Excellent)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = mean(Excellent), color = "blue", linetype = 2)
summary(Excellent)
```

```{r Very Good Gen Health and Sleep}
VG <- subset(brfss2013, brfss2013$genhlth == "Very good")
VG <- VG$sleptim1[!(is.na(VG$sleptim1))]
VG <- subset(VG, VG > 0 & VG < 24)
ggplot() + aes(VG)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = mean(VG), color = "blue", linetype = 2)
summary(VG)
```

```{r Good Gen Health and Sleep}
Good <- subset(brfss2013, brfss2013$genhlth == "Good")
Good <- Good$sleptim1[!(is.na(Good$sleptim1))]
Good <- subset(Good, Good > 0 & Good < 24)
ggplot() + aes(Good)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = mean(Good), color = "blue", linetype = 2)
summary(Good)
```

```{r Fair Gen Health and Sleep}
Fair <- subset(brfss2013, brfss2013$genhlth == "Fair")
Fair <- Fair$sleptim1[!(is.na(Fair$sleptim1))]
Fair <- subset(Fair, Fair > 0 & Fair < 24)
ggplot() + aes(Fair)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = mean(Fair), color = "blue", linetype = 2)
summary(Fair)
```

```{r Poor Gen Health and Sleep}
Poor <- subset(brfss2013, brfss2013$genhlth == "Poor")
Poor <- Poor$sleptim1[!(is.na(Poor$sleptim1))]
Poor <- subset(Poor, Poor > 0 & Poor < 24)
ggplot() + aes(Poor)+ geom_histogram(binwidth=1, colour="black", fill="white") + geom_vline(xintercept = mean(Poor), color = "blue", linetype = 2)
summary(Poor)
```

We can see above that there is a clear relationship between the mean amount of sleep reported and general health reported. While the reported numbers are discrete, as the sleep is only recorded in whole hour integers, we can even see the median move from 7 hours, for the 4 "most healthy" categorical variables, to 6 hours. 

**Research quesion 2:**

Now that we have observed a clear decrease in sleep as reported general health decreases, let's look into the relationship between sleep and exercise. To do this, I will split the exercise (exeroft1) data into 8 evenly split groups and compare the amoung of sleep reported for each group.

```{r Exercise Processing}
exercise <- brfss2013[!(is.na(brfss2013$exeroft1)), ]
exercise <- exercise[!(is.na(exercise$sleptim1)), ]
# Plot histrogram to get a feel for the data.
ggplot(exercise, aes(x = exeroft1)) + geom_histogram(bins = 100)
# The data clusters at a little over 100 and a little over 200 times walking.
# Below we split the data into 8 different groups in order to campare different levels of exercise and amount of sleep reported.
quantile(exercise$exeroft1, c(1/8, 1/4, 3/8, 1/2, 5/8, 3/4, 7/8))
ex1 <- subset(exercise, exercise$exeroft1 <= 102)
ex2 <- subset(exercise, exercise$exeroft1 > 102 & exercise$exeroft1 <= 103)
ex3 <- subset(exercise, exercise$exeroft1 > 103 & exercise$exeroft1 <= 104)
ex4 <- subset(exercise, exercise$exeroft1 > 104 & exercise$exeroft1 <= 105)
ex5 <- subset(exercise, exercise$exeroft1 > 105 & exercise$exeroft1 <= 107)
ex6 <- subset(exercise, exercise$exeroft1 > 107 & exercise$exeroft1 <= 203)
ex7 <- subset(exercise, exercise$exeroft1 > 203 & exercise$exeroft1 <= 215)
ex8 <- subset(exercise, exercise$exeroft1 > 215)
```

Now that we have 8 groups divided by the amount of exercise reported, we will look at the summary of each group's sleep data and also plot the average amount of sleep per exercise group on a single plot.

```{r ExSlp Summaries and Plot}
summary(ex1$sleptim1)
summary(ex2$sleptim1)
summary(ex3$sleptim1)
summary(ex4$sleptim1)
summary(ex5$sleptim1)
summary(ex6$sleptim1)
summary(ex7$sleptim1)
summary(ex8$sleptim1)
slpex <- data.frame(Ex.Group = c("exgrp1", "exgrp2", "exgrp3", "exgrp4", "exgrp5", "exgrp6", "exgrp7", "exgrp8"), Avg.Slp = c(mean(ex1$sleptim1), mean(ex2$sleptim1), mean(ex3$sleptim1), mean(ex4$sleptim1), mean(ex5$sleptim1), mean(ex6$sleptim1), mean(ex7$sleptim1), mean(ex8$sleptim1)))
ggplot(slpex, aes(x = Ex.Group, y = Avg.Slp, group = 1)) + 
  geom_line() + 
  geom_point() + 
  xlab("Exercise Group") + 
  ylab("Average Hours of Sleep") + 
  ggtitle("Sleep vs Exercise") + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  geom_hline(yintercept = mean(exercise$sleptim1), linetype = "dashed", color = "blue")
```

Interestingly, there is no obvious relationship between the amount of sleep someone gets and the amount of exercise they perform. If you were to split the data into halves, the group permornig less exercise would have a much higher average sleep amount, but there is not clear linear relationship between amount of sleep and exercise.

**Research quesion 3:**

Below is a brief look at the relationship between income level and general health. First we will take a look at the income data, then plot the income data and general health data together.

```{r Income Data Summary}
Income <- brfss2013$income2
Income <- Income[!is.na(Income)]
summary(Income)
```

We can see through this summary the different levels of Income that have been designated. The categories are each assigned a numerical value below, where the lowest income level is 1 and the highest is 8.

```{r Income Histo}
Income <- as.numeric(Income)
hist(Income)
```

We can see that there are more and more data points as we jump higher and higher within the income levels, and that the highest level of income has the most amount of datapoints.

If we plot the the General Health data alongside the Income data, we can get a feel for how the two relate.

```{r Income and Health}
IncHlth <- select(brfss2013, income2, genhlth)
IncHlth <- IncHlth[!is.na(IncHlth$income2), ]
IncHlth <- IncHlth[!is.na(IncHlth$genhlth), ]
IncHlth$income2 <- as.numeric(IncHlth$income2)
IncHlth$genhlth <- as.numeric(IncHlth$genhlth)
# Create groups by General Health and see how Income level is affected.
Excellent <- subset(IncHlth, IncHlth$genhlth == 1)
VG <- subset(IncHlth, IncHlth$genhlth == 2)
Good <- subset(IncHlth, IncHlth$genhlth == 3)
Fair <- subset(IncHlth, IncHlth$genhlth == 4)
Poor <- subset(IncHlth, IncHlth$genhlth == 5)
IncHlth <- data.frame(General.Health = c("Excellent", "Very good", "Good", "Fair", "Poor"), Avg.Income.Level = c(mean(Excellent$income2), mean(VG$income2), mean(Good$income2), mean(Fair$income2), mean(Poor$income2)))
# Below we plot the relationship between income level and general healthy.
ggplot(IncHlth, aes(x = reorder(General.Health, Avg.Income.Level, sum), y = Avg.Income.Level, group = 1)) +
  geom_point() +
  geom_line() +
  ylab("Income Category (1 = Lowest Income Level)") +
  xlab("General Health Category") +
  ggtitle("Income vs General Health") +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The above plot shows that there is a very strong relationship between the reported general healthy and income level in this data. With a slope close to one, moving one categorical level in Income typically correlates to a full categorical jump of General Health in the same direction. This data points to a strong positive correlation between the two variables. 
